Effects of epidemiological structure on the transient evolution of HIV virulence

### 2016-06-24 ms/plos_latex.tex
added methods section back
more makefile/ms tweaks
working on paper

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

@@

\end{flushleft}
% Please keep the abstract below 300 words
% current is 298!
\section*{Abstract}
The evolutionary dynamics of parasite virulence over the
  course of an emerging epidemic have important implications both for
  our basic understanding of epidemiological dynamics and,
  potentially, for the outcomes of public health interventions. In
  general changes in fitness landscapes over the course of an epidemic
  will select for higher virulence during the early,
  exponential-growth phase of the epidemic, but quantitative outcomes
  can depend sensitively on biological details and the structure of
  mathematical models used to capture them.  Fraser, Shirreff, and
  co-workers have proposed a series of models for eco-evolutionary
  dynamics of HIV that are relatively detailed in their portrayal of
  the tradeoffs between transmission and virulence (mediated by
  set-point viral load, SPVL) and their heritability between
  hosts. However, these models use very simple implicit
  representations of the transmission process that ignore the
  partnership dynamics that previous research has found to be critical
  in predicting epidemics of sexually transmitted diseases.  We
  explore models that combine HIV virulence tradeoffs with a range of
  epidemiological structures, modeling partnership formation and
  dissolution and allowing for individuals to transmit disease outside
  of partnerships. We assess summary statistics such as the peak value
  of virulence (SPVL) and the time at which the peak occurs across all
  models and across a Latin hypercube sample that captures a realistic
  range of partnership dynamic parameters for sub-Saharan Africa. In
  order to account for the different interpretations of parameters
  across model structures, we scale all parameter sets to constrain
  the simulated epidemic growth rate to be identical, matching a
  realistic baseline value. Our primary result is that, for this
  particular model setting, the simplest random-mixing structure is
  actually the best approximation to the most realistic model; this
  surprising outcome occurs because the dominance of extra-pair
  contact in the realistic model tends to mask the effects of
  partnership structure.

@@

Pathogens such as HIV can evolve rapidly in response to changing
conditions; such changes include both the increases in disease
incidence over the course of the epidemic and treatment
interventions. While researchers have successfully used computational
models to explore these evolutionary dynamics, these models often
neglect epidemiological processes, such as the formation and
dissolution of sexual partnerships; other research has shown that
these processes can strongly affect epidemic outcomes. We built and
compared models with a range of realism and complexity in the way they
modeled both partnership dynamics and the possibility of sexual
contact outside of partnerships. In contrast to models of intermediate
complexity, simple random-mixing models with no partnership structure
at all actually gave the closest match to the most realistic and
complex models; extra-pair contact tended to wash out the effects
of epidemiological structure. Nevertheless, the large differences
in evolutionary dynamics among different epidemiological models
suggests that researchers trying to predict the evolution of
pathogens should proceed with caution.

@@

The evolution of pathogen virulence is a fundamental process in
evolutionary biology, of both theoretical and (potentially) practical
importance. The trade-off theory \cite{Ebert1999} --- which
postulates that parasite virulence can be explained as the long-term
evolutionary outcome of a saturating relationship between parasite
clearance rate and transmission rate --- has been criticized
\cite{EbertBull2003,alizon_adaptive_2015}, but has also been
successfully applied in a variety of host-pathogen systems \cite{Dwyer+1990,mackinnon1999genetic,jensen2006empirical,deroode2008virulence}. One
particularly interesting application of these ideas is the work by
Fraser \etal\ showing that HIV appears to satisfy the prerequisites of
the tradeoff theory: in a study of discordant couples (i.e. long-term
sexual partnerships with one infected and one uninfected partner), HIV
virulence as measured by the rate of progression to AIDS was both
heritable and covaried with the set-point viral load (i.e., the
characteristic virus load measured in blood during the intermediate
stage of infection), which in turn predicted the probability of
within-couple transmission
\cite{Fraser+2007,fraser_virulence_2014}. Subsequent studies
\cite{shirreff_transmission_2011,herbeck_hiv_2014} used these data to
parameterize mechanistic models of HIV virulence evolution, suggesting
that HIV invading a novel population would initially evolve increased
virulence, peaking after approximately [XXX] years and then declining
slightly to a long-stable virulence level.

The work of Shirreff \etal\ \cite{shirreff_transmission_2011}, and particularly the predicted
transient peak in HIV virulence midway through the epidemic,
highlights the importance of interactions between epidemiological and
evolutionary factors \cite{day_virulence_2004,alizon_price_2009}.
However, despite the attention to mechanistic detail at the individual
or physiological level, the epidemiological structures used in these
models are relatively simple.

@@

We add complexity to the epidemiological model following the general approach of Champredon \etal\ \cite{champredon_hiv_2013}; individuals join and leave partnerships at a specified rate, and can have sexual contact both within and outside of established partnerships. At the same time, our analysis somewhat simplifies the models of Shirreff \etal\ \cite{shirreff_transmission_2011}, for computational tractability; we check that our qualitative results are not sensitive to these simplifications. In order to explore how virulence evolution depends on epidemiological structure, we consider a series of models with increasing levels of complexity. In order to avoid dependence of the results on a particular set of parameters --- as we explain below, finding matching sets of parameters across models with widely differing epidemiological structures is challenging --- we evaluate our models across a wide range of parameters, again following Champredon \etal\ \cite{champredon_hiv_2013} in using a Latin hypercube design. For each model run, we compute a set of metrics (peak virulence, timing of virulence peak, equilibrium virulence) that summarize the evolutionary trajectory of a simulated HIV epidemic.

@@

@@

Like Shirreff \etal\ \cite{shirreff_transmission_2011}, we focus on the evolution of mean $\log_{10}$ set-point viral load, SPVL (which we denote as $\alpha$), rather than virulence (i.e. rate of progression to AIDS) itself. 
In contrast to Shirreff \etal, we use a single-stage disease model instead of accounting explicitly for progression through the three main stages of HIV infection (primary, asymptomatic, and disease), and we use a simple exponentially distributed infectious period instead of a more realistic Weibull-distributed infectious period. We account for varying transmission rates and durations of each disease stage by summing the durations of three stages (again based on Shirreff \etal's model) and taking the duration-weighted average of transmission rates of three stages. Thus the within-couple transmission rate, $\beta$, for our models is given by:
\begin{equation}
\beta (\alpha) = \frac{D_P \beta_P + D_A (\alpha) \beta_A (\alpha) + D_D \beta_D}{D_P + D_A (\alpha) + D_D},
\end{equation}
where the duration of infection ($D_P$ and $D_D$) and rate of transmission ($\beta_P$ and $\beta_D$) of the \textbf{P}rimary and \textbf{D}isease stages
of infection are independent of the host's SPVL. Following Shirreff \etal, the duration of infection ($D_A$) and rate of transmission ($\beta_A$) for the \textbf{A}symptomatic stage are Hill functions of the SPVL:

\begin{equation}
\begin{split}
D_A(\alpha) &= \frac{\tsub{D}{max} D_{50} ^{D_k}}{V_\alpha ^{D_k} + D_{50}^{D_k}}, \\
\beta_A(\alpha) &= \frac{\tsub{\beta}{max} V_\alpha ^ {\beta_k}}%
{V_\alpha^{\beta_k} + \beta_{50} ^{\beta_k}},
\end{split}
\end{equation}
where $V_{\alpha} = 10^\alpha$. 
The \textbf{u}ncoupled and \textbf{e}xtra-couple transmission rates are scaled by multiplying the \textbf{w}ithin-couple transmission rate $\beta$ by the contact ratios $c_u/c_w$ and $c_e/c_w$.

@@

Like Shirreff \etal\ \cite{shirreff_transmission_2011} we incorporate a between-host mutation process in the SPVL, but simplify Shirreff \etal's evolutionary model slightly by using a one-to-one genotype-phenotype mapping. 
The mutational process in our model is directly taken from Shirreff \etal. Over the course of infection, mutation occurs within the host. However, it is assumed that SPVL of an infected individual is determined by the SPVL at the time of infection for simplicity (and is not further affected by within-host mutation). Instead, the mutational effect takes place when an infected individual transmits the virus to a susceptible individual. First, the distribution of \Lspvl\ is discretized into a vector:
\begin{equation}
\alpha_i =  (\tsub{\alpha}{max} - \tsub{\alpha}{min})\frac{(i-1)}{n-1} + \tsub{\alpha}{min} \qquad i = 1,2,3, \dots n.
\end{equation}
We have experimented with varying degrees of discretization in the strain distribution (i.e., values of $n$); in our model runs comparing results with Shirreff \etal\ \cite{shirreff_transmission_2011} (Fig.~\ref{fig:panel3}) we use $n=51$ (i.e. a bin width of 0.1 \Lspvl\ for $\alpha$), but we find only small differences when reducing $n$ to 21 (bin width 0.25 \Lspvl), which we use for all other simulations.

@@

We developed six multi-strain evolutionary models, designed to cover a gamut between Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simplistic \cite{shirreff_transmission_2011} epidemiological structures, each of which is based on different assumptions regarding contact structure and partnership dynamics. Specifically, we focus on the effects of the assumptions of (1) instantaneous vs. non-instantaneous partnership formation and (2) zero vs. positive extra-partnership sexual contact and transmission on the evolution of mean \Lspvl.

Our first four models consider explicit partnership dynamics and are based on Champredon \etal's model \cite{champredon_hiv_2013}. Models 1 and 2 assume non-instantaneous partnership formation (i.e. individuals spend some time uncoupled, outside of partnerships) and consist of five states that are classified by infection status and partnership status. $S$ is the number of single (uncoupled) susceptible individuals, and $I$ is the number of single infected individuals. $SS$ is the number of susceptible-susceptible couples, $SI$ is the number of serodiscordant (susceptible-infected) couples, and $II$ is the number of concordant positive (infected-infected) couples. Model 1 includes extra-partnership contact (with both uncoupled individuals and individuals in other partnerships) whereas model 2 only considers within-couple transmission. Models 3 and 4 assume instantaneous partnership formation and thus consist of only the three partnered states: $SS$, $SI$, and $II$. Parallel to model 1 and 2, model 3 includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals don't exist in this model) and model 4 only considers within-couple transmission.

In contrast, models 5 and 6 are not explicitly structured.  Model 5 is an implicit serial monogamy model based on the epidemiological model used by Shirreff \etal\ \cite{shirreff_transmission_2011}. It is actually a random mixing model that consist of only two states, $S$ and $I$, and does not consider explicit partnership dynamics. However, to simulate the effect of instantaneous partnership formation, it uses an adjusted transmission rate that is derived from approximated basic reproduction number of a serial monogamy model \cite{hollingsworth_hiv1_2008}. Finally, model 6 is a simple random-mixing model.

The base model (i.e. model 1) for the first four models is an extension of Champredon \etal's model \cite{champredon_hiv_2013}. Individuals in single compartment acquire a partner at a rate, $\rho$, and partnerships dissolve at a rate, $c$. Infected individuals in a discordant partnership infect susceptible partner at a rate $\beta$ (within-couple transmission rate) and susceptible individuals outside the partnership at a rate $c_e$ (extra-couple transmission rate). Likewise, a single infected individual can infect any susceptible individuals at a rate $c_u$ through uncoupled mixing. Extra-couple and uncoupled transmission are modeled in a same way as Champredon \etal's model. All the details have been adapted to a multi-strain scenario. Model 2, 3, and 4 are derived from the base model by removing epidemiological details (partnership formation and uncoupled/extra-couple contact). Model details are explained in the appendix.

@@

Despite considerable effort by many researchers \cite{hollingsworth_hiv1_2008,champredon_hiv_2013}, the parameters determining the rate and structure of sexual partnership change and contact are still very uncertain; this led Champredon \etal\ \cite{champredon_hiv_2013} to adopt a Latin hypercube sampling (LHS) strategy \cite{blower_drugs_1991} that evaluates model outcomes over a range of parameter values. In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$). We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.

Latin hypercube sampling is done as in Champredon \etal\ \cite{champredon_hiv_2013}. For each parameter, $z$, its range is divided into $N = 1000$ equal intervals on a log scale:

\begin{equation}
z_i = \exp\left(\log(\tsub{z}{min}) + [\log(\tsub{z}{max}) - \log(\tsub{z}{min})] \frac{i-1}{N-1}\right) \qquad i = 1,...,N.
\end{equation}
For simplicity, we assume that all parameters are uniformly distributed on the log scale. Following the vectorization of a parameter range, a matrix is constructed so that each column contains a vector of a parameter series which it represents ($z_1$,...$z_N$). Then, each column is replaced with a random permutation series of itself. Now, each row contains a different parameter set that is used for each simulation run.

@@

\begin{table}[h!]
\caption{Parameter ranges/values.  Note that $c$ and $\rho$ values are doubled from Champredon \etal\  because we keep track of individuals, while they keep track of couples. Starred (*) parameters (used in Figure~\ref{fig:panel3}) are from \cite{shirreff_transmission_2011}.}
\centering
\begin{tabular}{c p{2in} c l}
\hline 
Notation & Description & Range/Value & Source\\
\hline % inserts single horizontal line
$\rho$ & Partnership formation rate & 1/10-2/5 & \cite{champredon_hiv_2013} \\
$c$ & Partnership dissolution rate & 1/15-1/5 (1.25*) & \cite{champredon_hiv_2013} \\
$c_u/c_w$ & Relative contact rate for uncoupled transmission & 1/5-5 & Assumption \\
$c_e/c_w$ & Relative contact rate extra-couple & 0.01-1 & \cite{champredon_hiv_2013} \\
$\beta_P$ & Rate of transmission during primary infection & 1.31-5.09 (2.76*) & \cite{hollingsworth_hiv1_2008} \\
$\beta_D$ & Rate of transmission during high transmission disease stage & 0.413-1.28 (0.76*) & \cite{hollingsworth_hiv1_2008} \\
$D_P$ & Duration of primary infection & 1.23/12-6/12 (0.25*) & \cite{hollingsworth_hiv1_2008,shirreff_transmission_2011} \\
$D_D$ & Duration of high transmission disease stage & 4.81/12-14/12 (0.75*) & \cite{hollingsworth_hiv1_2008} \\
$\tsub{\beta}{max}$ & Maximum rate of transmission during asymptomatic stage & 0.317 & \cite{shirreff_transmission_2011} \\
$\beta_{50}$ & SPVL at which infectiousness is half maximum & 13938 & \cite{shirreff_transmission_2011} \\
$\beta_k$ & Hill coefficient: steepness of increase in infectiousness as a function of SPVL & 1.02 & \cite{shirreff_transmission_2011} \\
$\tsub{D}{max}$ & Duration of primary infection & 25.4 & \cite{shirreff_transmission_2011} \\
$D_{50}$ & SPVL at which duration of asymptomatic infection is half maximum & 3058 & \cite{shirreff_transmission_2011} \\
$D_{k}$ & Hill coefficient: steepness of decrease in duration as a function of SPVL & 0.41 & \cite{shirreff_transmission_2011} \\
$\sigma_M$ & Mutation standard deviation of $\log_{10}$ SPVL & 0.12 & \cite{shirreff_transmission_2011} \\
$\tsub{\alpha}{min}$ & Minimum $\log_{10}$ SPVL & 2 & \cite{shirreff_transmission_2011}\\
$\tsub{\alpha}{max}$ & Maximum $\log_{10}$ SPVL & 7 & \cite{shirreff_transmission_2011}\\
$n$ & Number of strains & 21 (51*) & Assumption\\
\hline
\end{tabular}
\label{table:parmsTable}
\end{table}

The one completely new parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (Appendix); we need to pick a reasonable range for it. Champredon \etal\ \cite{champredon_hiv_2013} assume that the effective within-couple contact rate and effective uncoupled contact rate have the same range of 0.05 - 0.25.  Given Champredon \etal's parameter range, the possible maximum and minimum values of $c_u/c_w$ are 5 and 1/5. Therefore, we use 1/5-5 as the range for the parameter $c_u/c_w$. Although this adds more uncertainty to the parameter $c_u$ --- Champredon \etal's range implies a 5-fold difference whereas ours gives a 25-fold difference --- as there is not much known about the uncoupled transmission rate, we consider the wider range appropriate.

@@

One of the most difficult parts of model comparison is finding parameter sets that are commensurate against many different model structures. For the most part, our models are too complex to easily derive analytical correspondences among them. Given a numerical criterion, such as $r$ (initial exponential growth rate) or ${\cal R}_0$ (intrinsic reproductive number), we can adjust one or more parameters by brute force to ensure that all of the models match according to that criterion. While ${\cal R}_0$ is often considered the most fundamental property of an epidemic, and might thus seem to be a natural goal, here we focus on matching the initial growth rate $r$ for several reasons. First, our primary interest is in the transient evolutionary dynamics of virulence, which are more strongly affected by $r$ than ${\cal R}_0$. Second, $r$ is in general more directly observable in real epidemics; $r$ can be estimated by simply fitting an exponential curve to the initial incidence or prevalence curves \cite{ma_estimating_2014}, while ${\cal R}_0$ typically requires either (1) knowledge of \emph{all} epidemic parameters or (2) relatively sophisticated back-calculation based on $r$ and knowledge of the serial interval or generation interval of the disease \cite{wallinga_how_2007}. Thus, we scale a parameter so that every run has the same initial exponential growth rate in the incidence.

>>>>

% Remove comment for double spacing
\usepackage{setspace} 
\doublespacing

@@

\end{flushleft}
% Please keep the abstract below 300 words
% current is 298!
\section*{Abstract}
The evolutionary dynamics of parasite virulence over the
  course of an emerging epidemic have important implications both for
  our basic understanding of epidemiological dynamics and,
  potentially, for the outcomes of public health interventions.
  Changes in the fitness landscape
  will generally select for higher virulence during the early
  phase of an epidemic, but quantitative outcomes
  can depend sensitively on biological details and the structure of
  mathematical models used to capture them.  Fraser, Shirreff, and
  co-workers have proposed a series of models for eco-evolutionary
  dynamics of HIV that are relatively detailed in their portrayal of
  the tradeoffs between transmission and virulence (mediated by
  set-point viral load, SPVL) and their heritability between
  hosts. However, these models use implicit
  representations of the transmission process that drastically simplify the
  partnership dynamics that previous research has found to be critical
  in driving epidemics of sexually transmitted diseases.  We
  explore models that combine HIV virulence tradeoffs with a range of
  epidemiological structures, modeling partnership formation and
  dissolution and allowing for individuals to transmit disease outside
  of partnerships. We assess summary statistics such as the peak value
  of virulence (SPVL) and the time at which the peak occurs across all
  models and across a Latin hypercube sample that captures a realistic
  range of partnership dynamic parameters for sub-Saharan Africa. In
  order to account for the different interpretations of parameters
  across model structures, we scale all parameter sets to constrain
  the simulated epidemic growth rate to be identical, matching a
  realistic baseline value. For this
  particular model setting, the simplest random-mixing structure is
  actually the best approximation to the most realistic model; this
  surprising outcome occurs because the dominance of extra-pair
  contact in the realistic model tends to mask the effects of
  partnership structure.

@@

Pathogens such as HIV can evolve rapidly in response to changes
in their environments; such changes include both increases in disease
prevalence over the course of the epidemic and treatment
interventions. While researchers have successfully used computational
models to explore these evolutionary dynamics, these models often
neglect details such as the formation and
dissolution of sexual partnerships; other research has shown that
these processes can strongly affect epidemic outcomes. We built and
compared models that used different methods to
model both partnership dynamics and sexual
contact outside of partnerships. Models of intermediate
complexity predicted much lower peak virulence (15 years to
progress to AIDS) compared to both more realistic models
and simple random-mixing models with no partnership structure
at all (both approx. 7.25 years to progress to AIDS); 
extra-pair contact tended to wash out the effects
of epidemiological structure. The large differences
in evolutionary dynamics among different epidemiological models
suggests that researchers trying to predict the evolution of
pathogens should proceed with caution.

@@

The evolution of pathogen virulence is a fundamental process in
evolutionary biology, of both theoretical and (potentially) practical
importance. The trade-off theory \cite{Ebert1999} --- which
postulates that parasite virulence can be explained as the long-term
evolutionary outcome of a saturating relationship between parasite
clearance rate and transmission rate --- has been criticized
\cite{EbertBull2003,alizon_adaptive_2015}, but has also been
successfully applied in a variety of host-pathogen systems \cite{Dwyer+1990,mackinnon1999genetic,jensen2006empirical,deroode2008virulence}. One
particularly interesting application of these ideas is the work by
Fraser \etal\ showing that HIV appears to satisfy the prerequisites of
the tradeoff theory: in studies of discordant couples (i.e. long-term
sexual partnerships with one infected and one uninfected partner), HIV
virulence as measured by the rate of progression to AIDS was both
heritable and covaried with the set-point viral load (i.e., the
characteristic virus load measured in blood during the intermediate
stage of infection), which in turn predicted the probability of
transmission
\cite{Fraser+2007,fraser_virulence_2014}. Subsequent studies
\cite{shirreff_transmission_2011,herbeck_hiv_2014} used these data to
parameterize mechanistic models of HIV virulence evolution, suggesting
that HIV invading a novel population would initially evolve increased
virulence, peaking after approximately 100-200 years and then declining
slightly to a long-stable virulence level.

The work of Shirreff \etal\ \cite{shirreff_transmission_2011}, and particularly the predicted
transient peak in HIV virulence midway through the epidemic,
highlights the importance of interactions between epidemiological and
evolutionary factors \cite{day_virulence_2004,alizon_price_2009}.
However, despite these studies' attention to detail at the individual
or physiological level, the epidemiological structures used in these
models are relatively simple.

@@

We add complexity to the epidemiological model following the general approach of Champredon \etal\ \cite{champredon_hiv_2013}; individuals join and leave partnerships at a specified rate, and can have sexual contact both within and outside of established partnerships. In order to explore how virulence evolution depends on epidemiological structure, we consider a series of models with increasing levels of complexity. In order to avoid dependence of the results on a particular set of parameters --- as we explain below, finding matching sets of parameters across models with widely differing epidemiological structures is challenging --- we evaluate our models across a wide range of parameters, again following Champredon \etal\ \cite{champredon_hiv_2013} in using a Latin hypercube design. For each model run, we compute a set of metrics (peak virulence, timing of virulence peak, equilibrium virulence) that summarize the evolutionary trajectory of a simulated HIV epidemic.

@@

\section*{Materials and Methods}

@@

Like Shirreff \etal\ \cite{shirreff_transmission_2011}, we focus on the evolution of mean $\log_{10}$ set-point viral load, SPVL (which we denote as $\alpha$), rather than th rate of progression to AIDS itself
(we refer to SPVL as ``virulence'' hereafter).
In contrast to Shirreff \etal, we use a single-stage disease model instead of accounting explicitly for progression through the three main stages of HIV infection (primary, asymptomatic, and disease), and we use a simple exponentially distributed infectious period instead of a more realistic Weibull-distributed infectious period. We account for varying transmission rates and durations of each disease stage by summing the durations of three stages (again based on Shirreff \etal's model) and taking the duration-weighted average of transmission rates of three stages. Thus the within-couple transmission rate, $\beta$, for our models is given by:
\begin{equation}
\beta (\alpha) = \frac{D_P \beta_P + D_A (\alpha) \beta_A (\alpha) + D_D \beta_D}{D_P + D_A (\alpha) + D_D},
\end{equation}
where the duration of infection ($D_P$ and $D_D$) and rate of transmission ($\beta_P$ and $\beta_D$) of the Primary and Disease stages
of infection are independent of the host's SPVL. Following Shirreff \etal, the duration of infection ($D_A$) and rate of transmission ($\beta_A$) for the Asymptomatic stage are Hill functions of the SPVL:

\begin{equation}
\begin{split}
D_A(\alpha) &= \frac{\tsub{D}{max} D_{50} ^{D_k}}{V_\alpha ^{D_k} + D_{50}^{D_k}}, \\
\beta_A(\alpha) &= \frac{\tsub{\beta}{max} V_\alpha ^ {\beta_k}}%
{V_\alpha^{\beta_k} + \beta_{50} ^{\beta_k}},
\end{split}
\end{equation}
where $V_{\alpha} = 10^\alpha$. 
The uncoupled and extra-couple transmission rates are scaled by multiplying the within-couple transmission rate $\beta$ by the contact ratios $c_u/c_w$ and $c_e/c_w$.

@@

Like Shirreff \etal\ \cite{shirreff_transmission_2011} we incorporate a between-host mutation process in the SPVL, but simplify Shirreff \etal's evolutionary model by using a one-to-one genotype-phenotype mapping. 
The mutational process in our model is directly taken from Shirreff \etal. Over the course of infection, mutation occurs within the host. However, it is assumed that SPVL of an infected individual is determined by the SPVL at the time of infection (and is not further affected by within-host mutation). Instead, the mutational effect takes place when an infected individual transmits the virus to a susceptible individual. First, the distribution of \Lspvl\ is discretized into a vector:
\begin{equation}
\alpha_i =  (\tsub{\alpha}{max} - \tsub{\alpha}{min})\frac{(i-1)}{n-1} + \tsub{\alpha}{min} \qquad i = 1,2,3, \dots n.
\end{equation}
We have experimented with varying degrees of discretization in the strain distribution (i.e., values of $n$); in our model runs comparing results with Shirreff \etal\ \cite{shirreff_transmission_2011} (Fig.~\ref{fig:panel3}) we use $n=51$ (i.e. a bin width of 0.1 \Lspvl\ for $\alpha$), but we find only small differences when reducing $n$ to 21 (bin width = 0.25 \Lspvl), which we use for all other simulations.

@@

We developed six multi-strain evolutionary models covering a gamut between Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} epidemiological structures, each of which is based on different assumptions regarding contact structure and partnership dynamics. Specifically, we focus on the effects of the assumptions of (1) instantaneous vs. non-instantaneous partnership formation and (2) zero vs. positive extra-partnership sexual contact and transmission on the evolution of mean \Lspvl.

Our first four models consider explicit partnership dynamics and are based on Champredon \etal's model \cite{champredon_hiv_2013}. The first two (``pair-formation'' or ``pairform'' for short) assume non-instantaneous partnership formation (i.e. individuals spend some time uncoupled, outside of partnerships) and consist of five states that are classified by infection status and partnership status. $S$ is the number of single (uncoupled) susceptible individuals, and $I$ is the number of single infected individuals. $SS$ is the number of concordant negative (susceptible-susceptible) couples, $SI$ is the number of serodiscordant (susceptible-infected) couples, and $II$ is the number of concordant positive (infected-infected) couples. The first (``pairform+epc'') includes extra-partnership contact (with both uncoupled individuals and individuals in other partnerships) whereas the second (``pairform'') only considers within-couple transmission. The next two models assume instantaneous partnership formation (``instswitch'') and thus consist of only the three partnered states: $SS$, $SI$, and $II$. Like the first two models, these models differ in their inclusion of extra-pair contact: the third model (``instswitch+epc'') includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals don't exist in this model) and the fourth (``instswitch'') only considers within-couple transmission.

The last two models do not explicitly track sexual partnerships.  One (``implicit'') is an implicit serial monogamy model based on the epidemiological model used by Shirreff \etal\ \cite{shirreff_transmission_2011}. It is actually a random mixing model that consist of only two states, $S$ and $I$, and does not consider explicit partnership dynamics. However, to reflect the effect of (instantaneously formed) partnership structure, it uses an adjusted transmission rate that is derived from an approximation of the basic reproduction number of a serial monogamy model \cite{hollingsworth_hiv1_2008}. Finally, the last model (``random'') is a simple random-mixing model.

The base model (i.e. pairform+epc) for the first four models is an extension of Champredon \etal's model \cite{champredon_hiv_2013}. Individuals in single compartment acquire a partner at a rate $\rho$, and partnerships dissolve at a rate $c$. Infected individuals in a discordant partnership infect their susceptible partner at a rate $\beta$ (within-couple transmission rate) and susceptible individuals outside the partnership at a rate $c_e$ (extra-couple transmission rate). Likewise, a single infected individual can infect any susceptible individuals at a rate $c_u$ through uncoupled mixing. Extra-couple and uncoupled transmission are modeled in the same way as Champredon \etal's model. All the details have been adapted to a multi-strain scenario. The second through fourth models (pairform, instswitch+epc, instswitch) are derived from the base model by simplifying epidemiological processes (partnership formation and uncoupled/extra-couple contact). Model details are explained in the appendix.

@@

Despite considerable effort \cite{hollingsworth_hiv1_2008,champredon_hiv_2013}, the parameters determining the rate and structure of sexual partnership change and contact are still very uncertain; this led Champredon \etal\ \cite{champredon_hiv_2013} to adopt a Latin hypercube sampling (LHS) strategy \cite{blower_drugs_1991} that evaluates model outcomes over a range of parameter values. In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the early- and late-stage transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$). We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.

Latin hypercube sampling is done as in Champredon \etal\ \cite{champredon_hiv_2013}. For each parameter, $z$, its range is divided into $N = 1000$ equal intervals on a log scale:
\begin{equation}
z_i = \exp\left(\log(\tsub{z}{min}) + [\log(\tsub{z}{max}) - \log(\tsub{z}{min})] \frac{i-1}{N-1}\right) \qquad i = 1,...,N.
\end{equation}
Random permutations of these vectors form columns in a sample parameter matrix; each row contains a different parameter set that is used for one simulation run.

@@

\begin{table}[h!]
\caption{Parameter ranges/values.  Note that $c$ and $\rho$ values are doubled from Champredon \etal\  because we keep track of individuals, while they keep track of couples. Starred (*) parameters (used in Figure~\ref{fig:panel3}), and Descriptions of Hill function coefficients, are taken from \cite{shirreff_transmission_2011}.}
\centering
\begin{tabular}{c p{2in} c l}
\hline 
Notation & Description & Range/Value & Source\\
\hline % inserts single horizontal line
$\rho$ & Partnership formation rate & 1/10-2/5 & \cite{champredon_hiv_2013} \\
$c$ & Partnership dissolution rate & 1/15-1/5 (1.25*) & \cite{champredon_hiv_2013} \\
$c_u/c_w$ & Relative contact rate for uncoupled transmission & 1/5-5 & Assumption \\
$c_e/c_w$ & Relative contact rate extra-couple & 0.01-1 & \cite{champredon_hiv_2013} \\
$\beta_P$ & Rate of transmission during primary infection & 1.31-5.09 (2.76*) & \cite{hollingsworth_hiv1_2008} \\
$\beta_D$ & Rate of transmission during high transmission disease stage & 0.413-1.28 (0.76*) & \cite{hollingsworth_hiv1_2008} \\
$D_P$ & Duration of primary infection & 1.23/12-6/12 (0.25*) & \cite{hollingsworth_hiv1_2008,shirreff_transmission_2011} \\
$D_D$ & Duration of high transmission disease stage & 4.81/12-14/12 (0.75*) & \cite{hollingsworth_hiv1_2008} \\
$\tsub{\beta}{max}$ & Maximum rate of transmission during asymptomatic stage & 0.317 & \cite{shirreff_transmission_2011} \\
$\beta_{50}$ & SPVL at which infectiousness is half maximum & 13938 & \cite{shirreff_transmission_2011} \\
$\beta_k$ & Hill coefficient: steepness of increase in infectiousness as a function of SPVL & 1.02 & \cite{shirreff_transmission_2011} \\
$\tsub{D}{max}$ & Duration of primary infection & 25.4 & \cite{shirreff_transmission_2011} \\
$D_{50}$ & SPVL at which duration of asymptomatic infection is half maximum & 3058 & \cite{shirreff_transmission_2011} \\
$D_{k}$ & Hill coefficient: steepness of decrease in duration as a function of SPVL & 0.41 & \cite{shirreff_transmission_2011} \\
$\sigma_M$ & Mutation standard deviation of $\log_{10}$ SPVL & 0.12 & \cite{shirreff_transmission_2011} \\
$\tsub{\alpha}{min}$ & Minimum $\log_{10}$ SPVL & 2 & \cite{shirreff_transmission_2011}\\
$\tsub{\alpha}{max}$ & Maximum $\log_{10}$ SPVL & 7 & \cite{shirreff_transmission_2011}\\
$n$ & Number of strains & 21 (51*) & Assumption\\
\hline
\end{tabular}
\label{table:parmsTable}
\end{table}

The one completely new parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (Appendix); we need to pick a reasonable range for it. Champredon \etal\ \cite{champredon_hiv_2013} assume that the effective within-couple contact rate and effective uncoupled contact rate have the same range of 0.05 - 0.25.  Given Champredon \etal's parameter range, the possible maximum and minimum values of $c_u/c_w$ are 5 and 1/5. Therefore, we use 1/5-5 as the range for the parameter $c_u/c_w$. Although this adds more uncertainty to the parameter $c_u$ --- Champredon \etal's range implies a 5-fold difference whereas ours gives a 25-fold difference --- we consider the wider range appropriate, as little is not much known about the uncoupled transmission rate.

@@

One of the most difficult parts of model comparison is finding
parameter sets that are commensurate with many different model
structures. For the most part, our models are too complex to easily
derive analytical correspondences among them. Given a numerical
criterion, such as $r$ (initial exponential growth rate) or ${\cal
  R}_0$ (intrinsic reproductive number), we can adjust one or more
parameters by brute force to ensure that all of the models match
according to that criterion. While ${\cal R}_0$ is often considered
the most fundamental property of an epidemic, and might thus seem to
be a natural matching criterion, here we focus on matching the initial
growth rate $r$ for several reasons. First, our primary interest is in
the transient evolutionary dynamics of virulence, which are more
strongly affected by $r$ than ${\cal R}_0$. Second, $r$ is 
more directly observable in real epidemics; $r$ can be estimated by
fitting an exponential curve to the initial incidence or
prevalence curves \cite{ma_estimating_2014}, while ${\cal R}_0$
typically requires either (1) knowledge of \emph{all} epidemic
parameters or (2) calculations based on
$r$ and knowledge of the serial interval or generation interval of the
disease \cite{wallinga_how_2007}. Thus, we scale parameters so that
every run has the same initial exponential growth rate of incidence.
